Train Harder, Smarter:\\Using Graph Theory to Create Route Suggestions

### 2013-11-18 paper.tex
Did a lot of work on improving the code and added Brent's algorithm to the writeup
Brent's algorithm to be implemented in the code shortly.

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage{algpseudocode}
\usepackage[toc,page]{appendix}
\usepackage{url}

@@

Once you recall that our distance goal, $d$, is the circumfrence of the loop,
it is easy to see that the radius is $\frac{d}{2\pi}$. We can then adjust our
equation to correspond with our starting point and radius as follows:
\[ x = \frac{d}{2\pi}(\cos ( t + \phi ) - \cos \phi ) + x_0.x \]
\[ y = \frac{d}{2\pi}(\cos ( t + \phi ) - \cos \phi ) + x_0.y \]
Of course, this needs some adjustment to deal with the difference of using
latitude and longitude over a cartesian coordinate system, since latitude
and longitude are not actually coordinates on a 2-d plane, but rather angles
that are made against the center of the Earth. For now, this is sufficient to
get an idea of what is happening, but we'll cover how this will be converted
into a latitude/longitude format in Section~\ref{sec:latlong}.

Until now, we've used $t$ to represent the angle in the parametric equation,
without delving into what it really means. However, we can use a much more
useful expression to represent this angle: $\frac{2\pi}{d}d_i$, where $d_i$ is
the current distance travelled, gives us the
angle in terms of a percentage of the requisite distance travelled. This gives
us the advantage of knowing exactly where in the loop we should be at any
current distance travelled. This also needs to be capped after $d_i \geq d$.
Incorporating this yields the final equations:
\[ x = \frac{d}{2\pi}(\cos (\min(\frac{2\pi}{d}d_i,2\pi) + \phi) - \cos \phi) + x_0.x \]
\[ y = \frac{d}{2\pi}(\sin (\min(\frac{2\pi}{d}d_i,2\pi) + \phi) - \sin \phi)+ x_0.y. \]

@@

The next step is to utilize these equations to calculate $\theta$. At any point
along the algorithm, we are at some $x_i$, and we have travelled a distance,
$d_i$. Figure~\ref{fig:algvals} should help to understand what exactly the
algorithm is attempting to achieve. The idea is this: if we have travelled some
distance $d_i$, then there is an $x_{\mbox{{\tiny ideal}}}$ on the ``ideal''
loop. The arc between $x_{\mbox{{\tiny ideal}}}$ and $x_0$ will be of length
$d_i$. The algorithm searches for a point $x_j$ such that the arc between
$x_{\mbox{{\tiny ideal}}}$ and $x_j$ will have the \emph{same} distance as the
line between $x_i$ and $x_j$. Fortunately, we have framed the equations for
calculating $x_j$ in terms of the distance travelled along the arc, and the
distance between $x_i$ and $x_j$ can easily be calculated using the pythagorean
theorem. Note that we find $x_j$ as follows:
\begin{align}
  x_j.x = \frac{d}{2\pi} ( & \cos( \frac{ 2 \pi }{ d } ( \min( d, d_i + d_j ) )
                                  + \phi ) \notag \\ & - \cos \phi ) + x_0.x
                                  \label{eq:xjx} \\
  x_j.y = \frac{d}{2\pi} ( & \sin( \frac{ 2 \pi }{ d } ( \min( d, d_i + d_j ) )
                                  + \phi ) \notag \\ & - \sin \phi ) + x_0.y.
                                  \label{eq:xjy}
\end{align}
To satisfy our constraint, we have
\begin{equation}
  d_j = \sqrt{ ( x_j.x - x_i.x )^2 + ( x_j.y - x_i.y )^2 }. \label{eq:dists}
\end{equation}
Obviously, this is a rather difficult equation to truly solve, so we must only
heuristically find a solution. Luckily, we
have a fairly small search space, since $d - d_i > d_j > 0$. On the other hand,
it is not readily apparent how these solutions will behave. We can, however,
be certain that it is continuous, so we can proceed as follows:
\begin{enumerate}
\item Check if the optimal solution would require that we move around the loop
  again. This will happen if
  \[d_i+\sqrt{(x_0.x-x_i.x)^2+(x_0.y-x_i.y)^2} > d.\]
  In this case, we set $x_j = x_0$, and stop searching.
\item \emph{NEEDS MORE WORK}
\end{enumerate}

At this point, we have found $x_j$. From here, calculating $\theta$ is fairly
straightforward:
\begin{align}
  \tan \theta = & \frac{x_j.y - x_i.y}{x_j.x-x_i.x} \notag \\
  \theta = & \tan^{-1} \frac{x_j.y - x_i.y}{x_j.x-x_i.x} \label{eq:theta}
\end{align}

Finally, we have a method for calculating the weights based on distance and
direction! Use \eqref{eq:xjx}, \eqref{eq:xjy}, and \eqref{eq:dists} to find
an $x_j$, and plug into \eqref{eq:theta} to find a $\theta$. Finally, use
$\theta$ in conjunction with \eqref{eq:weight} to assign weights to
directions.

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[toc,page]{appendix}
\usepackage{url}

@@

Once you recall that our distance goal, $d$, is the circumfrence of the loop,
it is easy to see that the radius is $\frac{d}{2\pi}$. We can then adjust our
equation to correspond with our starting point and radius as follows:
\[ x = \frac{d}{2\pi}(\cos ( t + \phi ) - \cos \phi ) + x_0 \]
\[ y = \frac{d}{2\pi}(\cos ( t + \phi ) - \cos \phi ) + y_0 \]
Of course, this needs some adjustment to deal with the difference of using
latitude and longitude over a cartesian coordinate system, since latitude
and longitude are not actually coordinates on a 2-d plane, but rather angles
that are made against the center of the Earth. For now, this is sufficient to
get an idea of what is happening, but we'll cover how this will be converted
into a latitude/longitude format in Section~\ref{sec:latlong}.

Until now, we've used $t$ to represent the angle in the parametric equation,
without delving into what it really means. However, we can use a much more
useful expression to represent this angle: $\frac{2\pi}{d}d_i$, where $d_i$ is
the current distance travelled, gives us the
angle in terms of a percentage of the requisite distance travelled. This gives
us the advantage of knowing exactly where in the loop we should be at any
current distance travelled. This also needs to be capped after $d_i \geq d$.
Incorporating this yields the final equations:
\[ x = \frac{d}{2\pi}(\cos (\min(\frac{2\pi}{d}d_i,2\pi) + \phi) - \cos \phi) + x_0 \]
\[ y = \frac{d}{2\pi}(\sin (\min(\frac{2\pi}{d}d_i,2\pi) + \phi) - \sin \phi)+ y_0. \]

@@

The next step is to utilize these equations to calculate $\theta$. At any point
along the algorithm, we are at some $x_i$, and we have travelled a distance,
$d_i$. Figure~\ref{fig:algvals} should help to understand what exactly the
algorithm is attempting to achieve. The idea is this: if we have travelled some
distance $d_i$, then there is an $x_{\mbox{{\tiny ideal}}}$ on the ``ideal''
loop. The arc between $x_{\mbox{{\tiny ideal}}}$ and $x_0$ will be of length
$d_i$. The algorithm searches for a point $x_j$ such that the arc between
$x_{\mbox{{\tiny ideal}}}$ and $x_j$ will have the \emph{same} distance as the
line between $x_i$ and $x_j$. Fortunately, we have framed the equations for
calculating $x_j$ in terms of the distance travelled along the arc, and the
distance between $x_i$ and $x_j$ can easily be calculated using the pythagorean
theorem. Note that we find $x_j$ as follows:
\begin{align}
  x_j = \frac{d}{2\pi} ( & \cos( \frac{ 2 \pi }{ d } ( \min( d, d_i + d_j ) )
                                  + \phi ) \notag \\ & - \cos \phi ) + x_0
                                  \label{eq:xj} \\
  y_j = \frac{d}{2\pi} ( & \sin( \frac{ 2 \pi }{ d } ( \min( d, d_i + d_j ) )
                                  + \phi ) \notag \\ & - \sin \phi ) + y_0.
                                  \label{eq:yj}
\end{align}
To satisfy our constraint, we have
\[ d_j = \sqrt{ ( x_j(d_j) - x_i )^2 + ( x_j(d_j) - x_i )^2 }. \]

\subsubsection{Finding $d_j$}
Obviously, this is a rather difficult equation to truly solve, so we must only
heuristically find a solution. Unfortunately, even the heuristic solution is
a difficult one for a few reasons. Most problematically, the sine and cosine
functions have up to two stationary points on the interval we are considering
and there may not be a solution of $d_j$ that will satisfy the equality.
Nonetheless, we will do our best to mitigate these issues and apply a good
heuristic to find $d_j$. First, we rearrange our equation so that we can apply
a root-finding approximation algorithm to it:
\begin{equation}
  f(d_j) = \sqrt{(x_j(d_j) - x_i)^2 + (y_j(d_j) - y_i)^2} - d_j. \label{eq:dj}
\end{equation}
This simply allows the $d_j$ we are looking for to be the zero of the function.
The first exciting bit of news is that for most solutions of this equation with
$d_j < d$, it will work out that $f(0) > 0$ and $f(d-d_i) < 0$.
We know that $f(0) \geq 0$ by the
positivity of norms, and the square root of the sum of the squares is the
euclidian norm. We also can be certain that $f(d-d_i) < 0$ since
the norm is also unable to change by more than the diameter
of the ideal loop, or even less based on the proximity of $x_i$ to the center of
the loop. In situations where $f(d-d_i) > 0$, we notice that the distance to the
start of the loop is further away than the amount of distance we have left to
travel; this means that we should just route back to the start anyways. As such,
we have bounds inside of which we know that there exists a solution. We apply
Brent's method, a robust root approximation algorithm, to determine the solution
inside of this zone. The idea behind Brent's method is to combine the bisection
method, the secant method, and inverse quadratic interpolation. Readers
unfamiliar with any of these techniques should refer to CITATIONS. Some
pseudocode is given for Brent's Method in Algorithm~\ref{alg:brent}. The idea
is this:

First, we need a bracket in which our function has a root. For us, this is
simple enough; we just use $[0, d-d_i]$. If that bracket does not satisfy the
condition, we just set $d_j = d-d_i$ and don't bother with anything.

The algorithm details out the actual swapping of $a_k$ and $b_k$, but here
we'll just assume that $|f(a_k)| > |f(b_k)|$ at all points in time; that is,
$b_k$ is a better approximation of the solution than $a_k$. We also require
$b_{k-1}$; for now, we'll set it to $a_k$.

This is a heuristic, so we'll keep repeating the algorithm until either
$a_k \approx b_k$ or $f(b_k) \approx 0$. At the $k^{\mbox{{\tiny th}}}$ iteration,
we'll calculate $b_{k+1}$ using both inverse quadratic interpolation and the
bisection method.

In general, the inverse quadratic interpolation can proceed very quickly, but we
add the bisection method to ensure convergence. In using interpolation, one of
the following conditions must be satisfied:
\begin{itemize}
\item Inverse Quadratic Interpolation was performed last time,
  \[ |s - b_k| < \left|\frac{b_{k-1}-b_{k-2}}{2}\right|, \]
  and
  \[|\delta| < |b_{k-1} - b_{k-2}| \]
\item The Bisection Method was used last time,
  \[|s - b_k| < \left|\frac{b_k-b_{k-1}}{2}\right|, \]
  and
  \[ |\delta| < |b_k - b_{k-1}| \]
\end{itemize}
The rest of the pseudocode serves only to ensure that variables are assigned
correctly. One notes that mflag indicates whether interpolation was used last
time or not.

\begin{algorithm}
  \caption{Using Brent's Method to find a zero of a function}
  \label{alg:brent}
  \begin{algorithmic}
    \Require $f(a_0)f(b_0) > 0$
    \Function{Brent}{$f : [a,b] \to \mathbb{R}$, $a_0$, $b_0$}
      \If{$|f(a_0)| < |f(b_0)|$}
        \Call{swap}{$a_0$,$b_0$}
      \EndIf
      \State $b_{k-1} \gets a_0$
      \State mflag $\gets$ True
      \While{$f(b_k) \not = 0$ {\bf and} $|b_k - a_k| > \epsilon$}
      \begin{align*}
        s \gets &\frac{a_kf(b_k)f(b_{k-1})}{(f(a_k)-f(b_k))(f(a_k)-f(b_{k-1}))} \\
        & + \frac{b_kf(a_k)f(b_{k-1})}{(f(b_k)-f(a_k))(f(b_k)-f(b_{k-1}))} \\
        & + \frac{b_{k-1}f(a_k)f(b_k)}{(f(b_{k-1})-f(a_k))(f(b_{k-1})-f(b_k))}
      \end{align*}
      \State {\bf if} $\displaystyle s \not \in \left[ \frac{3a_k+b_k}{4},b_k \right]$
             {\bf or}
             \State ~~~~(mflag is True {\bf and}
             \State ~~~~~~~~ ($\displaystyle |s-b_k| \geq \frac{|b_k-b_{k-1}|}{2}$
             \State ~~~~~~~~~~ {\bf or} $|b_k-b_{k-1}| < |\delta|$)) {\bf or}
             \State ~~~~(mflag is False {\bf and}
             \State ~~~~~~~~ ($\displaystyle |s-b_k| \geq \frac{|b_{k-1}-b_{k-2}|}{2}$
             \State ~~~~~~~~~~ {\bf or} $|b_{k-1}-b_{k-2}| < |\delta|$)) {\bf then}
             \State ~~~~ $\displaystyle s \gets \frac{a_k+b_k}{2}$
             \State ~~~~ mflag $\gets$ True
             \State {\bf else}
             \State ~~~~ mflag $\gets$ False
        \State $b_{k-2} \gets b_{k-1}$
        \State $b_{k-1} \gets b_k$
        \If{$f(a_k)f(s) < 0$}
          \State $b_k \gets s$
        \Else
          \State $a_k \gets s$
        \EndIf
        \If{$|f(a_k)| < |f(b_k)|$}
          \Call{swap}{$a_k$,$b_k$}
        \EndIf
      \EndWhile
    \EndFunction
  \end{algorithmic}
\end{algorithm}

Thus, by utilizing Brent's Algorithm, we can find $x_j$. From here, calculating
$\theta$ is fairly straightforward:
\begin{align}
  \tan \theta = & \frac{y_j - y_i}{x_j-x_i} \notag \\
  \theta = & \tan^{-1} \frac{y_j - y_i}{x_j-x_i} \label{eq:theta}
\end{align}

Finally, we have a method for calculating the weights based on distance and
direction! Use Brent's algorithm as described in Algorithm~\ref{alg:brent}
with the function to be zeroed as \eqref{eq:dj}. Use that $d_j$ in conjunction
with \eqref{eq:xj} and \eqref{eq:yj} to find
$x_j$ and $y_j$, respectively, and plug into \eqref{eq:theta} to find a
$\theta$. Finally, use $\theta$ in conjunction with \eqref{eq:weight} to assign
weights to directions.
